---
title: "Tutorial 7: Logistic Regression and Time Series Forecasting"
author: 'REPLACE WITH YOUR NAME and ID#'
date: 'Due by October 17, 9:00 AM'
output: html_document
---


## Submission Instructions

- Select `output: html_document`.
- Include all code chunks, so include `echo=TRUE` in all chunks.
- Replace the placeholder text, "Type your answer here.", with your own.
- Submit *only* the required question for grading (Part 2: Submission). You can delete everything else for that submission. Remember to include any `library('package_name')` statements that you'll need to run your code and future reproduction. 
- Rename your R Markdown file `T[X]_[MatricNumber].rmd`, and the output will automatically be `T[X]_[MatricNumber].html`. 
    - for example, `T6_12345.html`
    - X is the Tutorial number at the top of this file.
- Submit your both R Markdown file (.rmd) and HTML (.html) to Luminus for tutorial assignments (upload to Luminus under the correct Submission Folder). We shall do the same for practical exam.
- **It is important to be able to code and produce your Rmarkdown output file *independently*.** You are responsible for de-bugging and programming in the practical exam.

## Preparation

```{r load-libraries, echo=FALSE, warning = FALSE, message = FALSE}
# load required packages
# install any package below if it's first time loaded in your computer.
library(dplyr)
library(tidyr)
library(tseries) 
library(TTR) # One alternative for time-series in R
library(forecast) # An alternative for time series in R
library(car) # "Companion to Applied Regression" package, for F-test for linear combination of regression coefs
library(wooldridge) # wooldridge data set will be used in this tutorial
library(ggplot2) # optional. we expect you to know base graphics, but allow ggplot if you find it easier
```

## Part Two: Assignment Submission (25 points)

### Question 3 (total 17 points)

We will study fertility rates with two time-series datasets on U.S. and Singapore, respectively. 

High fertility rate is essential for long-term growth in any economy. Many countries are troubled with low or even negative fertility rate. For example, the fertility rate in Singapore in general shows a decreasing trend in recent decades (https://www.channelnewsasia.com/news/singapore/number-of-babies-born-in-singapore-falls-to-lowest-in-8-years-11743722). Fertility is affected by many socio-economic factors, including single rate, family disposable income, level of tax duty, war attrition, contraception technology, etc. 

Let's first take a look at the fertility rate in the United States in 20th century. `fertil3` data contains information about woman's fertility rate and personal tax exemption in U.S. in early-mid 1900s. 

- Dataset required: `data('fertil3')` in `wooldridge` package.

(Note: This dataset comes from a publicly available dataset from Jeffery Wooldridge Textbook. See data description in https://rdrr.io/cran/wooldridge/man/fertil3.html)

First, load in the time series data for this question. There are 72 observations on 24 variables about women fertility rate between year 1913 and 1984. Key variables are listed below:

- `gfr`: births per 1000 women between age 15 and 44.
- `pe`: real value personal tax exemption in US dollars.
- `pe_1`: 1-period lag of `pe`, real value personal tax exemption in US dollars from last year.
- `pe_2`: 2-period lag of `pe`, real value personal tax exemption in US dollars from two years ago.
- `ww2`: a binary variable = 1 during World War 2 between 1941 and 1945.
- `pill`: a binary variable = 1 from 1963 on when the birth control pill was made available for contraception.

```{r q3-read-dataset, echo=TRUE}
# read dataset into workplace, note that you need library(wooldridge) to load this data set
data('fertil3')
# if you want, is can be converted to ts object
fertil = ts(fertil3, frequency = 1, start = 1913)
```

Q3(a) Start off by plotting `gfr` (fertility rate) and `pe_1` (personal tax exemption from last year) against time in the same plot. What do you observe from the time series plots alone: do you see any trend or seasonality? Do you think that `gfr` time series is stationary? (2 points)

```{r q3a, echo=TRUE}
plot(fertil3$year, fertil3$gfr, type = 'l', col = 2, xlab = 'Year', ylab='Values', ylim = c(0,250))
lines(fertil3$year, fertil3$pe_1, col = 3)
legend("topright", c('gfr', 'pe_1'), lty = 1, col = 2:4)

x <- as.numeric(fertil3$gfr)
adf.test(x, alternative = c('stationary', 'explosive'))

```

Remark: You should be able to visualize data with R base graphics at least. Other alternative R graphic packages such as `ggplot` are welcomed additions.

<p style="color:red">**Type your answer here.**</p>
For 'pe_1',there is an increasing trend but no apparent seasonal effect. 
For 'gfr', there is a slight decreasing trend but no apparent seasonal effect. Due to the presence of trend, it is likely that the 'gfr' time series is not stationary. 

# (ignore) Upon conducting Dickey-Fuller test, we can observe that the p-value is large (0.623 > 0.05 at 5% significance level). Thus, we have insufficient evidence to reject the null hypothesis that the 'gfr' time series is not stationary. (we accept the alternative hypothesis that it is stationary) 

Q3(b) Many labor economists believe that fertility rate is affected by the economic policy such as personal tax exemption, e.g., `pe` and their lag terms. Run a linear regression of `gfr` on `pe`, `pe_1`, and `pe_2`. Interpret the coefficient before `pe`. Is it statistically significant? What's your conclusion about whether tax exemption improves fertility rate, from the regression output? (3 points)

```{r q3b, echo = TRUE}
summary(lm(gfr~pe + pe_1 + pe_2,fertil))
```

<p style="color:red">**Type your answer here.**</p>

Coeff of 'pe' = -0.01584. On average, as one-unit increase in the real value personal tax exemption in US dollars, the births per 1000 women btw age 15 and 44 decreases by 0.01584 units, keeping all other indepdent variables constant. As the p-value is large (0.910 >> 0.05 at 5% significance level), there is insufficient evidence to reject null hypothesis that the coefficient for 'pe' is 0. Thus, the coefficient is not signficantly different from 0 implying that 'pe' is not a statistically signficant predictor of 'gfr'.

(idk the conclusion part lol)



Q3(c) Now, include a time trend variable `t` into the regression model in part (b). Interpret the coefficients for `pe_2` and `t`. What's been changed in the regression result compared to previous one, in terms of coefficients (its sign and magnitude), goodness-of-fit R-square and F-test of the linear regression model? Which model do you think we should choose, and why? (4 points)


```{r q3c, echo = TRUE}
summary(lm(gfr~pe + pe_1 + pe_2 + t,fertil))
```

<p style="color:red">**Type your answer here.**</p>

Coefficient of 'pe_2' = 0.309055. On average, as one-unit increase in 'pe_2' results, the births per 1000 women btw age 15 and 44 increases by 0.309055. units, keeping all other independent variables constant. This coeff is higher than that in 3a. The p-value in this regression for 'pe_2' is also significantly smaller and <0.05 at 5% significance level. Thus, unlike in 3b, we have sufficient evidence to reject the null-hypothesis that the coefficient of 'pe_2' is 0 and 'pe_2' is a statistically significant predictor for 'gfr'. 

Coefficient of 't' = -1.068292. On average, as one-unit of time passes, the births per 1000 women btw age 15 and 44 decreases by 1.068292 units, keeping all other independent variables constant. 

In 3b, the R-squared value =  0.006135 < 0.5 and the F-test value = 0.1358 is small. 
In 3c, the R-squared value = 0.6229 > 0.5 and the F-test value = 26.84 is large. 
Thus, we should choose the model in 3c due to the higher R-squared value and F-test value as this implies that the regression statistics explained due to the model is higher than that due to residuals which means that the model in 3c has a higher predictive power. 


As we have discussed fertility rate in U.S. between 1913 and 1984 in previous question, now let's turn our attention to fertility rate in Singapore in modern times, where the fertility rate plummets to lowest figure in 2020 in history. 

- Dataset required: `SGfertil20.csv`

Note: This dataset comes from publicly available data from the Singapore Department of Statistics, or SingStat. https://www.singstat.gov.sg/find-data/search-by-theme/population/births-and-fertility/latest-data. You can download the data yourself and explore. For homework, I have done some data cleaning as to have only one time series variable ,`gfr`, in `sgfertil.csv` which contains information on fertility rate of childbearing age women between age 15 and 44 in Singapore comparable to that in U.S. 

```{r q3-fertil-data-input, echo=TRUE}
# read raw data from 'sgfertil.csv'
sgfertil = read.csv(file = 'SGfertil20.csv', header= TRUE)
colnames(sgfertil)[1]='Data.Series'

sgfertil = sgfertil %>% 
  # given our focus in univariate time-series analysis on total fertility rate..
  select(Data.Series, Total.Fertility.Rate..Per.Female.) %>%
  # rename
  rename(year = Data.Series, gfr = Total.Fertility.Rate..Per.Female.) %>%
  # sort the data bt ascending on year
  arrange(year)
# extract the key time series of our interest and convert it to ts obj
gfrsg = ts(sgfertil$gfr, start = 1960, end = 2020, frequency = 1)
```
`sgfertil` contains yearly data about gross fertility rate per thousand women between 1960 and 2020. A quick description of the data:

Obs:  61
- `gfr`: total fertility rate of women in their childbearing age between 15 and 44, #births per thousand
- `year`: as described.

Q3(d) First plot the `gfrsg` against time. Draw the time series plot of `gfrsg`. Describe what you observe from the plot: do you see any trend or/and seasonality? Do you think total fertility rate time series in Singapore `gfrsg` stationary? (2 points)

```{r q3d, echo = TRUE}
plot(gfrsg)

adf.test(gfrsg) 
```


<p style="color:red">**Type here your answer here.**</p>
There seems to be a decreasing trend. There does not seem to be any apparent seasonal effect. Due to the presence of trend, it is likely that the time series in 'gfrsg' is not stationary. 



Q3(e) In order to predict future fertility rate using information contained in the time series of `gfrsg` itself, let's make forecast by simply fitting a moving average series. First, compute the moving averages with window length of 4 years and plot both the original and moving average series in the same graph. Second, obtain the moving average prediction series. What is the MA predicted value for Year 2021? (3 points)

```{r qn3e, echo = TRUE}
ma_4yrs <- TTR :: SMA(gfrsg, n=16)
plot(ma_4yrs, col = 2, ylim = c(0,6), ylab = 'fertility rate')
lines(gfrsg, col = 3)
legend ("topright", c('moving average series', 'original series'), lty =1, col = 2:3)

##how to do ma prediction series 

```

<p style="color:red">**Type here your answer here.**</p>




Q3(f) Based on your observation in part (d), predict the fertility rates for Year 2021 and Year 2024 with Holt-Winter model of your choice. (3 points)

```{r qn3f}
hw <- HoltWinters(gfrsg, gamma=FALSE)
hw
predict(hw, n.ahead = 4)
predict(hw, n.ahead = 16)

```

<p style="color:red">**Type here your answer here.**</p>
For 2021, MA predicted value is 1.022681.
For 2024, MA predicted value is 0.7907255.

### Question 4 (Total 8 points)

- Dataset required: `data('recid')`

Recidivism rate in Singapore is 24% in 2016 (https://data.gov.sg/dataset/recidivism-rate). Criminals tend to relapse into criminal offense after the release from the prison. Recidivism is costly and causes serious social and economical problem. It is not only wasteful with the resources invested in prison, including staffing, infrastructure investment, daily operation cost, and economic opportunity cost for both prisoners and staffs (i.e., labor values that could be generated elsewhere other than being locked up and guarding in prison, respectively), but also harms the community for the second time due to crime re-commitment. Recidivism is thus a critical evaluating metric for prison performance, e.g., rehabilitation or training program. `recid.csv` contains 1445 observations of recidivism cases in United States where it currently has the largest prison population in the world (about one out every five peope imprisoned in the world is incarcerated in the United States).

```{r q3-data-input, echo=TRUE}
recid = read.csv(file = 'recid.csv', header= TRUE)
```

C.-F. Chung, P. Schmidt, and A.D. Witte (1991), “Survival Analysis: A Survey,” Journal of Quantitative Criminology 7, 59-98.
Data Source: https://www.cengage.com/cgi-wadsworth/course_products_wp.pl?fid=M20b&product_isbn_issn=9781111531041. A quick description of the data:

Obs:  1445

  1. black                    =1 if black
  2. alcohol                  =1 if alcohol problems
  3. drugs                    =1 if drug history
  4. super                    =1 if release supervised
  5. married                  =1 if married when incarc.
  6. felon                    =1 if felony sentence
  7. workprg                  =1 if in N.C. pris. work prg.
  8. property                 =1 if property crime
  9. person                   =1 if crime against person
 10. priors                   # prior convictions
 11. educ                     years of schooling
 12. rules                    # rules violations in prison
 13. age                      in months
 14. tserved                  time served, rounded to months
 15. follow                   length follow period, months
 16. durat                    max(time until return, follow) in month
 17. cens                     =1 if duration right censored
 18. ldurat                   log(durat)

Q4(a) Criminal offense lawyers usually fight for supervised release of the offender. Given the seven variables in `recid` data set: `super`, `rules`, `age`, `tserved`, `married`, `priors` and `black`, how could you help the law firm to predict the outcome of a recent case given that a 55-year-old married black client has been serving the jail time for 5 years and 11 months with no prior conviction and during which broken no rule in the prison? 
In order to evaluate the likelihood to answer such question. First of all, write down the regression model you propose to use. Clearly define or label your variables.  (2 point)

<p style="color:red">**Type your answer here**</p>

logit('super') = b0 + b1'age' + b2'black' + b3'married' + b4'priors' + b5'tserved' + b6'rules'
b0: Log-odds when all seven variables are 0. Baseline odds is exp(b0)
bx where x = 1 to 6: As one unit of variable increases, log-odds increases/decreases by bx, holidng all other constant. Or one unit of variable increase multiplies the baseline odds by exp(bx).


Q4(b) Run your regression model in R and interpret the coefficients before `priors` and `black`. Based on the results, do you think there was any racial discrimination in terms of likelihood of supervised release, and why? (3 point)
```{r qn4b, echo = TRUE}
summary(glm(super ~ age + black + married + priors + tserved + rules, family=binomial, data=recid))
```

<p style="color:red">**Type your answer here**</p>

b2 = 0.1114900. Being black increases the log-odds of supervised release by 0.1114900, holding all other constant.
b4 = -0.0943194. As the number of prior convictions increases by 1 unit, the log-odds of supervised release decreases by 0.0943194, holding all others constant.

Thus, we can conclude there is racial discrimination in favour of being black as being black increases the likelihood of supervised release by exp(0.1114900) = 1.117 times. 

Q4(c) Finally, what is the predicted probability for supervised release for the new client? (2 point)

<p style="color:red">**Type your answer here**</p>

log(p)= b0 + b1'age' + b2'black' + b3'married' + b4'priors' + b5'tserved' + b6'rules'

#idk what to code for this 

Q4(d) Successfully debug the Rmarkdown file and produce an HTML for submission. (1 point)


